---
title: "Assignment 4: Data Wrangling"
author: "Jess Garcia"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Tuesday, Feb 16 @ 11:59pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1
#Check working directory & load packages
getwd()

library(tidyverse)
library(lubridate)

#Upload & rename four raw data files
EPAair.O3.2018.data <- read.csv("C:/Users/93jes/Documents/ENV872/Environmental_Data_Analytics_2021/Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE)

EPAair.O3.2019.data <- read.csv("C:/Users/93jes/Documents/ENV872/Environmental_Data_Analytics_2021/Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE)

EPAair.PM25.2018.data <- read.csv("C:/Users/93jes/Documents/ENV872/Environmental_Data_Analytics_2021/Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE)

EPAair.PM25.2019.data <-read.csv("C:/Users/93jes/Documents/ENV872/Environmental_Data_Analytics_2021/Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE)


#2. Explore dimensions, column names, & structure of the datasets

dim(EPAair.O3.2018.data)
dim(EPAair.O3.2019.data)
dim(EPAair.PM25.2018.data)
dim(EPAair.PM25.2019.data)

colnames(EPAair.O3.2018.data)
colnames(EPAair.O3.2019.data)
colnames(EPAair.PM25.2018.data)
colnames(EPAair.PM25.2019.data)

str(EPAair.O3.2018.data)
str(EPAair.O3.2019.data)
str(EPAair.PM25.2018.data)
str(EPAair.PM25.2019.data)

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3. Change date to date

#EPAair.O3.2018.data
#EPAair.O3.2019.data
#EPAair.PM25.2018.data
#EPAair.PM25.2019.data

#Confirm class of date column is factor and not date variable
class(EPAair.O3.2018.data$Date)
class(EPAair.O3.2019.data$Date)
class(EPAair.PM25.2018.data$Date)
class(EPAair.PM25.2019.data$Date)

#Change date to date
EPAair.O3.2018.data$Date <-as.Date(EPAair.O3.2018.data$Date, 
                                   format = "%m/%d/%Y") 
EPAair.O3.2019.data$Date <-as.Date(EPAair.O3.2019.data$Date, 
                                   format = "%m/%d/%Y") 
EPAair.PM25.2018.data$Date <-as.Date(EPAair.PM25.2018.data$Date, 
                                     format = "%m/%d/%Y") 
EPAair.PM25.2019.data$Date <-as.Date(EPAair.PM25.2019.data$Date, 
                                     format = "%m/%d/%Y") 


#4. Select columns

EPAair.O3.2018.data.selected <- select(EPAair.O3.2018.data, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

EPAair.O3.2019.data.selected <- select(EPAair.O3.2019.data, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

EPAair.PM25.2018.data.selected <- select(EPAair.PM25.2018.data, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

EPAair.PM25.2019.data.selected <- select(EPAair.PM25.2019.data, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5. For PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5"

#view(EPAair.PM25.2018.data)
vignette("dplyr")
class(EPAair.PM25.2018.data.selected$AQS_PARAMETER_DESC)


EPAair.PM25.2018.data.selected2 <-EPAair.PM25.2018.data.selected%>%
  mutate(AQS_PARAMETER_DESC = "PM2.5")

EPAair.PM25.2019.data.selected2 <-EPAair.PM25.2018.data.selected%>% 
  mutate(AQS_PARAMETER_DESC = "PM2.5")


#6. Save processed datasets
write.csv(EPAair.O3.2018.data.selected, row.names = FALSE, file = "./Data/Processed/EPAair.O3.2018.data.selected_processed.csv")

write.csv(EPAair.O3.2019.data.selected, row.names = FALSE, file = "./Data/Processed/EPAair.03.2019.data.selected_processed.csv")

write.csv(EPAair.PM25.2018.data.selected2, row.names = FALSE, file = "./Data/Processed/EPAair.PM25.2018.data.selected2_processed.csv")

write.csv(EPAair.PM25.2019.data.selected2, row.names = FALSE, file = "./Data/Processed/EPAair.PM25.2019.data.selected2_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. ##use the datasets from #4 before you changed them
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels) #Don't have to use intersect for this, just use filter
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets (just the one) such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7. Combine four datasets with rbind
#Confirm column names are identical
#summary(EPAair.O3.2018.data.selected$Site.Name)
#summary(EPAair.O3.2019.data.selected$Site.Name)
#summary(EPAair.PM25.2018.data.selected$Site.Name)
#summary(EPAair.PM25.2019.data.selected2$Site.Name)


#Combine with rbind
EPAair_O3_PM25_NC1819_tight <- rbind(EPAair.O3.2018.data.selected,
    EPAair.O3.2019.data.selected, 
    EPAair.PM25.2018.data.selected2, 
    EPAair.PM25.2019.data.selected2)


#8. Wrangle new dataset
#view(EPAair_O3_PM25_NC1819_tight)
class(EPAair_O3_PM25_NC1819_tight$Date)

EPAair_O3_PM25_NC1819_tight.wrangled <- EPAair_O3_PM25_NC1819_tight %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory"| Site.Name == "Leggett" | Site.Name == "Hattie Avenue"| Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne"| Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School" & Site.Name!= "")%>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY)%>%
summarise(mean.AQI = mean(DAILY_AQI_VALUE), 
          mean.latitude = mean(SITE_LATITUDE),
          mean.longitude = mean(SITE_LONGITUDE))%>%
  mutate(Month = month(Date))%>%
  mutate(Year = year(Date))


#Confirm interim dimensions of dataset are: 14,752 x 9
dim(EPAair_O3_PM25_NC1819_tight.wrangled)
#Dimensions don't match. I have less rows than I'm supposed to, but not sure why. I can't figure it out, so I'm moving on


#9. Spread aka pivot_wider the dataset
EPAair_O3_PM25_NC1819_wide <- pivot_wider(EPAair_O3_PM25_NC1819_tight.wrangled, names_from = AQS_PARAMETER_DESC, values_from = mean.AQI)
#Somehow DAILY_AQI_VALUE disappeared as a column when I wrangled the data in question 8, so I guessed I'm supposed to used the newly created mean for AQI since I got confirmation that my code in question 8 was correct. 

                                            
#10. Dimensions of dataset
dim(EPAair_O3_PM25_NC1819_wide)


#11. Save new combined, tidy, processed dataset
write.csv(EPAair_O3_PM25_NC1819_wide, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a/b. Generate a summary data frame
summary.EPAair_O3_PM25_NC1819_wide <-(EPAair_O3_PM25_NC1819_wide)%>%
  group_by (Site.Name, Month, Year)%>%
  summarise(mean.ozone = mean(Ozone),
            mean.PM2.5 = mean(PM2.5))%>%
  drop_na(Month, Year)

#Try with na.omit instead of drop_na
summary2.EPAair_O3_PM25_NC1819_wide <-(EPAair_O3_PM25_NC1819_wide)%>%
  group_by (Site.Name, Month, Year)%>%
  summarise(mean.ozone = mean(Ozone),
            mean.PM2.5 = mean(PM2.5))%>%
  na.omit(Month, Year)

#13
dim(summary.EPAair_O3_PM25_NC1819_wide)
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: We used drop_na rather than na.omit because na.omit will actually remove any rows that contain NA (in month or year) from the dataset completely. Whereas drop_na will just omit or exclude NA from your values but keep the rows/observations that were excluded for having NAs.
