---
title: "Assignment 5: Data Visualization"
author: "Jess Garcia"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 23 at 11:59 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (both the tidy 
[`NTL-LTER_Lake_ChemistryPhysics_Nutrients_PeterPaul_Processed.csv`] and the gathered
[`NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv`] versions) and 
the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1. Confirm directory, load packages, & upload data

#Confirm workingdirectory
getwd()

#Load packages
#install.packages("cowplot")
library(tidyverse)
library(cowplot)
library(lubridate)

#Load data
ChemistryNutrients_PeterPaul <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

PeterPaul_Nutrients_Gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv", stringsAsFactors = TRUE)

Niwot_Littertrap <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)


#2. Confirm dates as date objects
class(ChemistryPhysics_PeterPaul$sampledate)
class(PeterPaul_Nutrients_Gathered$sampledate)
class(Niwot_Littertrap$collectDate)

#Change from factor objects to date objects

ChemistryNutrients_PeterPaul$sampledate <- as.Date(ChemistryNutrients_PeterPaul$sampledate, format = "%Y-%m-%d")

PeterPaul_Nutrients_Gathered$sampledate <- as.Date(PeterPaul_Nutrients_Gathered$sampledate, format = "%Y-%m-%d")

Niwot_Littertrap$collectDate <- as.Date(Niwot_Littertrap$collectDate, format = "%Y-%m-%d") 
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#https://ggplot2.tidyverse.org/reference/theme.html 

#Build default theme
JG_default_theme <- theme_bw(base_size = 12)+
  theme(axis.text = element_text(color = "black"), 
        plot.title = element_text(color = "black", size = 16, face = "bold",
        hjust = 0.5))

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
#4. Create scatterplot for phosphorous and phosphate at Pete & Paul Lakes
PeterPaul.tp_ug.po4.plot <-
  ggplot(ChemistryNutrients_PeterPaul, aes(x=po4, y = tp_ug))+
  geom_point(aes(color=lakename))+
  JG_default_theme+
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y")+
  ggtitle("Peter Lake & Paul Lake\nPhosphorous & Phosphate Levels")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(x = "Phosphate", y= "Phosphorous (ug)", color = "Lakes")+
  xlim(0, 50)+
  geom_smooth(method = lm)
print(PeterPaul.tp_ug.po4.plot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5. Create three separate boxplots 
#First make month as factor and not integer
ChemistryNutrients_PeterPaul$month <- as.factor(ChemistryNutrients_PeterPaul$month)

#Boxplot for temperature
PeterPaul_Temperature.boxplot <-
ggplot(ChemistryNutrients_PeterPaul, aes(x=month, y= temperature_C, color =lakename))+
  geom_boxplot()+
  ggtitle("Lake Temperature by Month")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(x= "Month", y = "Temp (C)", color = "Lakes")
print(PeterPaul_Temperature.boxplot)

#Boxplot for TP
PeterPaul_TP.boxplot <-
ggplot(ChemistryNutrients_PeterPaul, aes(x=month, y= tp_ug, color =lakename))+
  geom_boxplot()+
ggtitle("Lake Phosphorous Levels by Month")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(x= "Month", y = "Phosphorous (ug)", color = "Lakes")
print(PeterPaul_TP.boxplot)

#Boxplot for TN
PeterPaul_TN.boxplot <-
ggplot(ChemistryNutrients_PeterPaul, aes(x=month, y= tn_ug, color =lakename))+
  geom_boxplot()+
  ggtitle("Lake Nitrogen Levels by Month")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(x= "Month", y = "Nitrogen (ug)", color = "Lakes")
print(PeterPaul_TN.boxplot)

#Combine the threee boxplots using cowplot
PeterPaul_combined_boxplot <-
plot_grid(PeterPaul_Temperature.boxplot, PeterPaul_TP.boxplot +theme(legend.position="none"), PeterPaul_TN.boxplot +theme(legend.position ="none"), nrow=3, align = 'h')

print(PeterPaul_combined_boxplot)




PeterPaul_combined_boxplot <-
plot_grid(PeterPaul_Temperature.boxplot, PeterPaul_TP.boxplot, PeterPaul_TN.boxplot, nrow=3, align = 'h')

print(PeterPaul_combined_boxplot)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6. Niwot Ridge - Needles plot w/NLCD classes separated by color

Niwot_NeedlesMass_color.plot <- 
  ggplot(subset(Niwot_Littertrap, functionalGroup == "Needles"),
  aes(x=collectDate, y=dryMass, color = nlcdClass))+
  geom_point()+
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y")+
  theme(axis.text.x = element_text(angle=45, hjust=1.0))+
  ggtitle("Niwot Ridge: Needles collected in litter traps")+
  theme(plot.title = element_text(hjust = 0.5))+
  labs(x = "Collection Date", y= "Dry Mass (g)")

print(Niwot_NeedlesMass_color.plot)


#7. Niwot Ridge - Needles plot w/NLCD classes separated by facets

Niwot_NeedlesMass_facets.plot <-
  ggplot(subset(Niwot_Littertrap, functionalGroup == "Needles"),
  aes(x=collectDate, y=dryMass))+
  geom_point()+
  facet_wrap(vars(nlcdClass), nrow =3)+
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y")+
    theme(axis.text.x = element_text(angle=45, hjust=1.0))+
  ggtitle("Niwot Ridge: Needles collected in litter traps")+
    theme(plot.title = element_text(hjust = 0.5))+
  labs(x = "Collection Date", y= "Dry Mass (g)")

print(Niwot_NeedlesMass_facets.plot)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
